YALE DEPARTMENT OF ECONOMICS \\ \vspace*{60pt} The Folly of the Crowd: \\ Market Inefficiencies in Peer-to-Peer Lending

### 2015-03-31 draft.Rnw
Tinkered a bit. Especially with beginning.

\documentclass[12 pt, twoside]{article}
%\usepackage[doublespacing]{setspace}
\usepackage{graphicx, verbatim}
\usepackage[twoside, bindingoffset=5mm]{geometry}
\usepackage[font=small,labelfont=sf]{caption}
%\usepackage{chronosys}
\usepackage{float}
\usepackage{subfig}
\usepackage{url}
%%TIMELINE SHIT
\usepackage{array, booktabs} % for timeline
\usepackage[x11names]{xcolor}
\usepackage{colortbl}
\usepackage{caption}
\DeclareCaptionFont{blue}{\color{LightSteelBlue3}}

\newcommand{\foo}{\color{LightSteelBlue3}\makebox[0pt]{\textbullet}\hskip-0.5pt\vrule width 1pt\hspace{\labelsep}}
%%END
\setlength{\textwidth}{7in} 
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in} 
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\overfullrule=2cm
\usepackage{titlesec}
\usepackage[linktoc=all]{hyperref}
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black,urlcolor=black}
\usepackage[citestyle=verbose,bibstyle=numeric,sorting=ynt,backend=biber]{biblatex}
\usepackage[toc,page]{appendix}
\addbibresource{lit.bib}

\begin{document}
\setcounter{tocdepth}{2}
%\linespread{1.3}
\title{YALE DEPARTMENT OF ECONOMICS \\ \vspace*{60pt} The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending}
\author{Samuel Shleifer
  \footnote{Thanks to Professor Vineet Kumar for his help
  in developing the ideas and statistical methodologies involved in this paper,
  to Devin Pope and Rajkamal Iyer for help manipulating the Prosper data, 
  and to Andrei Shleifer, Nicholas Barberis and Andrew Metrick for helpful comments and suggestions.}
  \footnote{R code available at \url{github.com/sshleifer/ProsperThesis}}}
\date{\today}
\maketitle
\begin{center}
Essay submitted to the Faculty of the Department of Economics in partial fulfillment
of the requirements for the degree of Bachelor of Arts at Yale University.

Thesis Advisor: Vineet Kumar\footnote{Yale School of Management}
\end{center}

@@

\newpage
\begin{abstract}
This essay analyzes the effect of information gathering costs on financial market
efficiency, using data from Prosper.com, America's first Peer-to-Peer Lending Marketplace.
Peer-to-peer lending is the practice of individual and institutional investors
lending money to borrowers over the internet without the assistance of a
traditional financial intermediary. Between 2006 and 2011, Prosper underwent two
major policy changes that guided and then forced investors to lend at higher
interest rates, as well as to reduce their exposure to subprime borrowers.

During the policy regime before either policy intervention (2006-2008), investors set
interest rates by auction, made avoidably bad mistakes, and earned poor returns.
The data from this first regime is inconsistent with three of the predictions of
the Efficient Markets Hypothesis predictions: that (1) risky assets should earn positive expected returns,
(2) riskier assets should earn higher returns than less risky assets, and, most
conclusively, (3) ``prices should fully reflect publicly available information."

Controlling for macroeconomic differences between the time periods of each policy regime
reveals that inconsistencies (1) and (2), which disappeared after Prosper began
guiding investors towards setting higher interest rates, are partly the
consequence of a turbulent macroeconomic period. Inconsistency (3), that strategies
using publicly available information can earn superior risk-adjusted returns,
which holds until Prosper begins directly setting rates, suggests that 
centralized interest rate setting by a market maker may lead to a more efficient market than the
unconstrained interaction of supply and demand, especially when information
gathering costs are high.
\end{abstract}
\newpage
\tableofcontents

<<init,echo=FALSE,warning=FALSE,results='hide',message=FALSE,cache=FALSE,eval=TRUE>>=
#DONT CACHE SAYS YIHUI
require(dplyr)              ### Used Extensively
require(ggplot2)            ### Used Extensively
require(gridExtra)          ### For facet plots
require(knitr)              ### For chunk control, compilation
require(leaps)              ### For feature selection
require(lubridate)          ### For date grouping, conversions
require(reshape2)           ### For melt() before ggplot
require(scales)             ### For plots
require(texreg)             ### Regression Output using texreg()
require(tidyr)              ### For spread()
require(xtable)             ### Tabular Output using xtable()
options(scipen=999,digits=4,formatR.arrow=TRUE,width=90)
coruse = 'pairwise.complete.obs'
opts_chunk$set(list(fig.path='figure/minimal-', fig.align='center',echo=FALSE, fig.show='hold',tidy=TRUE,fig.height=5,fig.pos='H',warning=FALSE, message=FALSE,results='asis'))
@

@@

bigReg3.sc = regsubsets(def ~
                          BorrowerRate.sc + TotalCreditLinespast7years.sc + OpenRevolvingAccounts.sc + OpenRevolvingMonthlyPayment + InquiriesLast6Months + TotalInquiries + CurrentDelinquencies.sc + DelinquenciesLast7Years.sc + PublicRecordsLast10Years.sc + DebtToIncomeRatio.sc + StatedMonthlyIncome + TotalProsperLoans.sc + TotalProsperPaymentsBilled.sc + OnTimeProsperPayments.sc + ProsperPaymentsLessThanOneMonthLate.sc + ProsperPaymentsOneMonthPlusLate + ProsperPrincipalBorrowed.sc + MonthlyLoanPayment.sc +CreditScoreEst.sc + StatedMonthlyIncome.sc
                        ,nbest=1
                        ,nvmax=numvars
                        ,intercept=TRUE
                        ,really.big=TRUE
                        ,data = tr3)
@
%full train feat selection in Draft3.Rnw
<<testFuncs,results='hide',cache=FALSE>>=
makeForm<-function(bigReg, y = 'roi ~ grade +'){
  #FOR FEATURE SELECTION
  if(is.recursive(summary(bigReg))){
    wh = summary(bigReg)$which
    vars = names(wh[numvars,][wh[numvars,]==TRUE])[1:numvars+1]  
    form = paste(y, paste(vars,collapse=' + '), sep=' ')    
  }else{
    print('FAILING TO AUTOMATE FEATURE SELECTION')
    form = "roi ~ BorrowerAPR + OpenRevolvingAccounts + InquiriesLast6Months + TotalInquiries + CurrentDelinquencies + DelinquenciesLast7Years + MonthlyLoanPayment + CreditScoreEst"
  }
  return(form)
}

@@

ruleTest<-function(train){
  inq = quantile(train$TotalInquiries, probs=(.4), na.rm=TRUE)
  amt = quantile(train$LoanOriginalAmount, probs=(.4), na.rm=TRUE)
  best = subset(train, train$TotalInquiries < inq & train$LoanOriginalAmount < amt)
  return(best$roi)
}
meanUL<-function(vec){return(mean(unlist(vec),na.rm=TRUE))}
sdUL<-function(vec){return(sd(unlist(vec),na.rm=TRUE))}
@

\section{Introduction}
Historically, there are two major perspectives of efficient market prices. Fama
(1970)'s semi-strong efficient markets hypothesis ("EMH") predicts that a
strategy using publicly available information should not be able to earn excess
risk-adjusted returns. Risk models such as CAPM imply that risky assets should
have positive expected returns and that riskier assets should have higher
returns than less risky assets.\footcite{sharpe} The information aggregation perspective
associated with Hayek (1945) and Stiglitz and Grossman (1980), further
predicts that many individuals, each using their own information, collectively
arrive at a more efficient price than a central agent who sets prices, since the
agent cannot access all the information available to the crowd.\footcite{stiglitz}\footcite{hayek}

@@

A comparison of loan performance in different regimes allows me to consider
two contrasting views of how an internet marketplace such as Prosper works.
One view, closely linked to Hayek, suggests that the marketplace would work
best with minimal intervention from the market maker Prosper, so many
individual lenders incorporate their private knowledge and information in
the determination of the interest rate.   An opposing view would hold that
individuals are not sophisticated or knowledgeable enough to assess loan
risk, and some guidance or even interest rate setting from the market maker
Prosper would improve loan performance. The evidence on loan returns in the
three regimes points quite clearly in favor of the market maker hypothesis,
and against Hayek: as Prosper becomes more involved in the rate-setting process,
the excess returns available using publicly available information approach 0,
while aggregate investor returns exhibit higher mean and lower variance.

@@

Much of the evidence presented in this essay is inconsistent with an explanation
that all of the differences between regimes can be explained by differences in
time period. For example, in the late auction regime, when investors set
interest rates, strategies using publicly available information can earn excess
returns. In the Prosper-rate regime, when Prosper sets interest rates directly,
they cannot. The two periods are very similar macroeconomically. Perhaps most tellingly,
in the period when Prosper sets rates it seems to use precisely the information that was
helpful in selecting the best performing loans in the early auction regime.
The points strongly toward the conclusion that early investors were not using
publicly available information effectively.

@@

In October 2008, Prosper was shut down by the SEC because it was designated as
a seller of securities,\footnote{Rather than, as it argued, a marketplace
connecting buyers and sellers.} and re-opened in July 2009
after a 9 month quiet period with more stringent under-writing standards, and
greater guidance for lenders. Importantly, all the information about borrower
creditworthiness that is used in this paper is similar to what was available to
investors when they bid on loans. Since December 2010, the beginning of the
\textit{Prosper-rate} regime, rates have been determined by Prosper, on the
basis of a borrower's credit history.

\begin{table}
  %\renewcommand\arraystretch{1.4}\arrayrulecolor{LightSteelBlue3}
  %\captionsetup{singlelinecheck=false, font=blue, labelfont=sc, labelsep=quad}
  \caption{Timeline of Events}\vskip -1.5ex
  \begin{tabular}{@{\,}r <{\hskip 2pt} !{\foo}
    >{\raggedright\arraybackslash}p{12cm}}
    \toprule
    \addlinespace[1.5ex]
    Apr. 21, 2006 & First loan of early auction regime.\\
    Mar. 1, 2007 & Prosper bans borrowers with credit score $\leq$ 520.\\
    Oct. 15, 2008 & Early Auction period ends with SEC enforced shutdown.\\
    Jul. 20, 2009 & Prosper reopens with more informative credit grades, requires repeat borrowers to have credit score
    $\geq$ 600, and requires new borrowers to have credit score $\geq$ 640.
    Beginning of Early Auction Regime.\\
    Dec. 19, 2010 & Prosper-rate regime, where Prosper sets rates directly,
    begins, because "auctions made the deployment of funds more time consuming
    with little gain in lender returns,” - Prosper CEO Chris
    Larsen.\footcite{renton}\\
    Aug. 3, 2011 & Last loan in Prosper-rate regime originated.
  \end{tabular}
\end{table}[h]

Under all three regimes, lenders could choose to whom to lend and how much to
lend, allowing them to diversify across many loans, thereby decreasing the
variance of their returns. The minimum investment in a loan was \$25, with no
maximum, and lenders could hold as many loans as they wished.
Before a potential borrower's listing is posted on the website, Prosper authenticates their social security number, driver's
license and address, and pulls the borrower's credit history from Experian,
which includes credit score, total number of delinquencies, current
delinquencies, and many other historical credit variables.\footcite{freedman} 
In exchange for verifying borrowers' identities and managing the
marketplace, Prosper charges a 0.5-3 percentage
point origination fee on each loan depending on the borrower's risk profile and
loan duration. 
This fee,usually 1\%, is added to the borrower's monthly payment.
Lenders are also charged a servicing fee of 0.5-1\%. There are no prepayment penalties.
This rate is lower than that charged by most traditional financial institutions, which have the higher costs of physical branches, and pay for human analysis of credit history.\footcite{hanson}
The P2PL model relies on lenders being able to infer the riskiness of borrowers: since Prosper gets fees from monthly payments, they too are incentivized to maximize investor returns, both for short term profit, and long term reputation.\newline

In my sample, Prosper's origination volume, as shown in Figure \ref{fig:origTime}, peaked in the months before the shutdown, and then slowly resumed after the quiet period. Since 2011, when the sample ends, origination volume has eclipsed its 2008 peak.\newline
<<origTime, cache=FALSE, fig.cap='This chart shows two panels. The first panel shows the number of loans funded on Prosper.com in each month of the sample. The second panel shows the total dollars lent to borrowers on Propers.com in each month of the sample.'>>=
g2 = group_by(nd, mgroup, regime) %>%
  summarise(Number_Of_Loans = length(mgroup) 
            ,Dollars_Lent = sum(LoanOriginalAmount)) %>%
    melt(id.vars = c("mgroup",'regime'))

@@

\subsection{Auction Design}
From April 19, 2006 until December 19, 2010. Prosper operated a \emph{variable
rate} model, where lenders and borrowers determined interest rates using a Dutch
auction-like system, where each investor bid consisted of a (dollar amount,
interest rate) pair, such that the dollar amount is greater than \$50, the
minimum bid, and the rate is capped at the borrower's reservation
rate.\footcite{freedman}
In these auctions, borrowers would declare a maximum reservation interest
rate, capped at 36\%, where bidding would start. Over the course of a 7-10 day bidding period, that rate would be bid down by investors, with the loan's interest rate closing at the lowest rate $R_{clearing}$ such that enough investors were willing to fully fund the loan at or below the $R_{clearing}$. All investors who bid at or below the clearing rate would receive the clearing rate. If a borrower could not reach 100\% funding at his reservation rate, the loan would not be funded.\newline
As an instructional example, suppose Bill the borrower needed to borrow \$100, and set his maximum rate at 20\%.
Let Chris, David and Ellen be investors, with Chris bidding (\$50, 15\%), David bidding (\$50, 18\%) and Ellen bidding (\$30,19\%). In this scenario, Chris and David would win the auction, and each lend lend Bill \$50, and receive an 18\% interest rate. Bill would pay 18\% interest before fees.\newline

\subsection{Early Auction Regime}
In the early auction regime, loans originated between April 19, 2006 and October
15, 2008, auction participants had access to considerable information from
Prosper, including over 60  Borrower characteristics, such as residence,
employment status and an Experian credit score bucket, but few intelligent risk estimates. Additionally, Prosper did not regulate who could apply for a loan on the website. Besides basic fraud prevention, any applicant who received 100\% funding from investors was eligible, for Prosper's first year of operation. On March 1, 2007, amidst poor returns, Prosper announced that listings with credit scores below 520 would no longer be offered to investors. This intervention is not discussed for lack of space, but shows that after a year, more or less, Prosper felt the need to nudge a fully decentralized investment process in the right direction.

@@

\subsection{Prosper-rate Regime}
Under the current \emph{Prosper-rate} model, enacted December 20, 2010, lenders
no longer determine the loan rate in an auction. Instead, Prosper uses the
intelligent credit grades given to investors in the late auction regime to set
interest rates on loans, with slightly less stringent under-writing standards to
determine eligibility. The change was motivated by a desire to accelerate the
listing process, so that investors could reinvest money more quickly. Moreover,
towards the end of the late auction regime, auctions were arriving at interest
rates that were almost completely determined by the credit grade. Under the
fixed-rate regime, investors choose whether or not to invest in a given loan, at
the interest rate set by Prosper, so bids consist only of a dollar amount.\newline

@@

This essay continues Freedman and Jin's critical look at investor decisions in the early
auction regime, but with the benefit of three more years of data, and two more
major Prosper policy changes: informative grading and centralized rate-setting.
Instead of a focus on soft information, I describe the hard information that
went largely ignored by investors until Prosper aggregated it into more
informative credit grades. Instead of investor learning as the catalyst that led
to a more efficient market, I frame Prosper's rule changes as catalysts that
force investors to make better decisions.

Inconsistent with Hayek, the data suggest that when investors are left on their own
to set prices, markets are not nearly as efficient as the prices set by ``a
single mind'', namely a market maker who aggregates the data. Furthermore, as interest rate setting
becomes more centralized, interest rates are more sensitive to the hard credit variables
that actually predict default.

\section{Data}
My sample consists of 39,555 three year, unsecured (no collateral) Loans, worth
\$231.7 Million,  originated from 2006-2011. The data was downloaded off Prosper.com.
Borrowers apply for loans by creating a public listing on the Prosper.com website, and can request anywhere from \$1,000 to \$35,000.\footnote{The minimum increased to \$2,000 effective Nov. 1, 2011}.
Since our analysis requires measurement of loan performance, which can only take
place three  years after origination (given 3 year maturity on all loans), the
sample is ended in 2011, as we do not know what will end up happening to more recently originated loans. 
As shown in Table \ref{sample}, this leaves us with many more early auction regime loans than the post shutdown regimes.\newline
<<sampTab, cache=FALSE>>=
samp.tab = mf(count(nd,regime))
colnames(samp.tab) = c('Regime','n','Freq.')
xtable(samp.tab,kdigits=4, caption='Observations per regime. Each observation is
one funded loan.', label='sample')
@

@@

\subsection{Measuring Loan Performance with ROI}
<<roivstatus,cache=FALSE,fig.height=3,fig.cap="This chart shows the distribution of ROI for completed and defaulted loans over all three regimes. A continuous ROI variable offers more information about the quality of an investment than a binary default indicator. Dotted lines indicate mean ROI for category.",fig.height=3>>=
nd$d.factor =  ifelse(nd$defaulted == 1, 'Defaulted','Completed')
smeans = summarise(group_by(nd,d.factor), mRoi = mean(roi, na.rm=TRUE))
ggplot(nd, aes(x=roi,fill=d.factor)) +
  geom_histogram(binwidth=.01)+ 
  geom_vline(data=smeans, aes(xintercept=mRoi,color=d.factor),linetype='longdash') + 
  theme_bw() +
  labs(title='Distribution of ROI for each Loan Outcome (all regimes)', x='ROI',y='Count',fill='Loan Status') +
  scale_x_continuous(limits=c(-1, 1), breaks=seq(-1,1,by=.25),label=percent) +
  facet_wrap(~d.factor,ncol=1,scales='free_y')
  @

>>>>

\documentclass[12 pt, twoside]{article}
%\usepackage[doublespacing]{setspace}
\usepackage{graphicx, verbatim}
\usepackage[twoside, bindingoffset=5mm]{geometry}
\usepackage[font=small,labelfont=sf]{caption}
\usepackage{float}
\usepackage{subfig}
\usepackage{url}
%%TIMELINE SHIT
\usepackage{array, booktabs}
\usepackage[x11names]{xcolor}
\usepackage{colortbl}
\usepackage{caption}
\DeclareCaptionFont{blue}{\color{LightSteelBlue3}}
\newcommand{\foo}{\color{LightSteelBlue3}\makebox[0pt]{\textbullet}\hskip-0.5pt\vrule width 1pt\hspace{\labelsep}}
%%END

%BLANK SPACE BABY
\usepackage{afterpage}

\newcommand\blankpage{%
    \null
    \thispagestyle{empty}%
    \addtocounter{page}{-1}%
    \newpage}
%End

\setlength{\textwidth}{7in} 
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in} 
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\overfullrule=2cm
\usepackage{titlesec}
\usepackage[linktoc=all]{hyperref} %TOC Links
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black,urlcolor=black}
\usepackage[citestyle=verbose,bibstyle=numeric,sorting=ynt,backend=biber]{biblatex}
\usepackage[toc,page]{appendix}
\addbibresource{lit.bib}

\begin{document}
\setcounter{tocdepth}{2}

\title{YALE DEPARTMENT OF ECONOMICS \\ \vspace*{60pt} The Folly of the Crowd: \\
Market Inefficiencies in Peer-to-Peer Lending}
\author{Samuel Shleifer
  \footnote{Thanks to Professor Vineet Kumar for his help
  in developing the ideas and statistical methodologies involved in this paper,
  to Devin Pope and Rajkamal Iyer for help manipulating the Prosper data, 
  and to Andrei Shleifer, Nicholas Barberis and Andrew Metrick for helpful comments and suggestions.}
  \footnote{R code available at \url{github.com/sshleifer/ProsperThesis}}\\
Thesis Advisor: Vineet Kumar\footnote{Yale School of Management}}
\maketitle

\begin{center}
Essay submitted to the Faculty of the Department of Economics in partial fulfillment
of the requirements for the degree of Bachelor of Arts at Yale University.
\end{center}

@@

\afterpage{\blankpage}
\newpage
\begin{abstract}
This essay analyzes the effect of market maker interventions on financial market
efficiency, using data from Prosper.com, the first p2p lending
website in the US.\footnote{P2P lending is the practice of individual and
  institutional investors lending money to borrowers over the internet, without
the presence of a financial intermediary} Between 2006 and 2011, Prosper underwent two
major policy changes that guided and then forced investors to lend at higher
interest rates, as well as to reduce their exposure to subprime borrowers.

During the period before either change, investors set interest rates by auction,
made avoidably bad mistakes, and earned poor returns.
After Prosper began guiding investors towards setting higher interest rates,
they made fewer mistakes and returns improved.

This essay expands on the work of Freedman and Jin (2008), who use investor
learning to explain the improvement, with the benefit of three more years of
data and two major policy changes. 
Instead of investor learning as the catalyst that led to a more efficient market, I frame Prosper's rule changes as catalysts that guide
and then force investors towards making better decisions, and find that when
investors set rates by auction, investment strategies using publicly available
information can earn excess returns, an inconsistency with the predictions of
the Efficient Markets Hypothesis, but that once Prosper begins setting rates,
it is no longer possible to beat the market.
\end{abstract}
\newpage
\tableofcontents

<<init,echo=FALSE,warning=FALSE,results='hide',message=FALSE,cache=FALSE,eval=TRUE>>=
#DONT CACHE SAYS YIHUI
require(dplyr)              ### Used Extensively
require(ggplot2)            ### Used Extensively
require(gridExtra)          ### For facet plots
require(knitr)              ### For chunk control, compilation
require(leaps)              ### For feature selection
require(lubridate)          ### For date grouping, conversions
require(reshape2)           ### For melt() before ggplot
require(scales)             ### For plots
require(texreg)             ### Regression Output using texreg()
require(tidyr)              ### For spread()
require(xtable)             ### Tabular Output using xtable()
options(scipen=999,digits=4,formatR.arrow=TRUE,width=90)
coruse = 'pairwise.complete.obs'
opts_chunk$set(list(fig.path='figure/minimal-', fig.align='center',echo=FALSE,tidy=TRUE,fig.height=5,warning=FALSE, message=FALSE,results='asis',strip.white=TRUE))
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knit_hooks$set(document = hook2)
@

@@

bigReg3.sc = regsubsets(def ~
                          BorrowerRate.sc + TotalCreditLinespast7years.sc + OpenRevolvingAccounts.sc + OpenRevolvingMonthlyPayment + InquiriesLast6Months + TotalInquiries + CurrentDelinquencies.sc + DelinquenciesLast7Years.sc + PublicRecordsLast10Years.sc + DebtToIncomeRatio.sc + StatedMonthlyIncome + TotalProsperLoans.sc + TotalProsperPaymentsBilled.sc + OnTimeProsperPayments.sc + ProsperPaymentsLessThanOneMonthLate.sc + ProsperPaymentsOneMonthPlusLate + ProsperPrincipalBorrowed.sc + MonthlyLoanPayment.sc +CreditScoreEst.sc + StatedMonthlyIncome.sc
                        ,nbest=1
                        ,nvmax=numvars
                        ,intercept=TRUE
                        ,really.big=TRUE
                        ,data = tr3)
@

<<testFuncs,results='hide',cache=FALSE>>=
makeForm<-function(bigReg, y = 'roi ~ grade +'){
  #FOR FEATURE SELECTION
  if(is.recursive(summary(bigReg))){
    wh = summary(bigReg)$which
    vars = names(wh[numvars,][wh[numvars,]==TRUE])[1:numvars+1]  
    form = paste(y, paste(vars,collapse=' + '), sep=' ')    
  }else{
    print('FAILING TO AUTOMATE FEATURE SELECTION')
    form = "roi ~ BorrowerAPR + OpenRevolvingAccounts + InquiriesLast6Months + TotalInquiries + CurrentDelinquencies + DelinquenciesLast7Years + MonthlyLoanPayment + CreditScoreEst"
  }
  return(form)
}

@@

ruleTest<-function(train){
  inq = quantile(train$TotalInquiries, probs=(.4), na.rm=TRUE)
  amt = quantile(train$LoanOriginalAmount, probs=(.4), na.rm=TRUE)
  best = subset(train, train$TotalInquiries < inq & train$LoanOriginalAmount < amt)
  return(best$roi)
}
meanUL<-function(vec){return(mean(unlist(vec),na.rm=TRUE))}
sdUL<-function(vec){return(sd(unlist(vec),na.rm=TRUE))}
@
\section{Introduction}
Historically, there are two major perspectives on efficient market prices. Fama
(1970)'s semi-strong efficient markets hypothesis ("EMH") predicts that a
strategy using publicly available information should not be able to earn excess
risk-adjusted returns. Risk models such as CAPM imply that risky assets should
have positive expected returns and that riskier assets should have higher
returns than less risky assets.\footcite{sharpe} The information aggregation perspective
associated with Hayek (1945) and Stiglitz and Grossman (1980), further
predicts that many individuals, each using their own information, collectively
arrive at a more efficient price than a central agent who sets prices, since the
agent cannot access all the information available to the crowd.\footcite{stiglitz}\footcite{hayek}

@@

A comparison of loan performance in different regimes allows me to consider
two contrasting views of how an internet marketplace such as Prosper works.
One view, closely linked to Hayek, suggests that the marketplace would work
best with minimal intervention from the market maker Prosper, so many
individual lenders incorporate their private knowledge and information in
the determination of the interest rate.   An opposing view would hold that
individuals are not sophisticated or knowledgeable enough to assess loan
risk, and some guidance or even interest rate setting from Prosper, the market
maker, would improve loan performance. The evidence on loan returns in the
three regimes points quite clearly in favor of the market maker hypothesis
as Prosper becomes more involved in the rate-setting process,
the excess returns available using publicly available information approach 0,
while aggregate investor returns exhibit higher mean and lower variance.

@@

Much of the evidence presented in this essay is inconsistent with an explanation
that all of the differences between regimes can be explained by differences in
time period. For example, in the late auction regime, when investors set
interest rates, strategies using publicly available information can earn excess
returns. In the Prosper-rate regime, when Prosper sets interest rates directly,
they cannot. The two periods are very similar macroeconomically. Perhaps most tellingly,
in the period when Prosper sets rates it seems to use similar variables to those
that were helpful in selecting the best performing loans in the early auction
regime, pointing strongly toward the conclusion that early investors were not using
publicly available information effectively.

@@

In October 2008, Prosper was shut down by the SEC because it was designated as
a seller of securities,\footnote{Rather than, as it argued, a marketplace
connecting buyers and sellers.} and re-opened in July 2009
after a 9 month quiet period with more stringent under-writing standards, and
greater guidance for lenders. Importantly, all the information about borrower
creditworthiness that is used in this paper is similar to what was available to
investors when they bid on loans. Since December 2010, the beginning of the
\textit{Prosper-rate} regime, rates have been determined by Prosper, on the
basis of a borrower's credit history, as provided by Experian.

\begin{table}
  %\renewcommand\arraystretch{1.4}\arrayrulecolor{LightSteelBlue3}
  %\captionsetup{singlelinecheck=false, font=blue, labelfont=sc, labelsep=quad}
  \caption{Timeline of Events}\vskip -1.5ex
  \begin{tabular}{@{\,}r <{\hskip 2pt} !{\foo}
    >{\raggedright\arraybackslash}p{12cm}}
    \toprule
    \addlinespace[1.5ex]
    Apr. 21, 2006 & First loan of early auction regime.\\
    Mar. 1, 2007 & Prosper bans borrowers with credit score $\leq$ 520.\\
    Oct. 15, 2008 & Early Auction period ends with SEC enforced shutdown.\\
    Jul. 20, 2009 & Prosper reopens with more informative credit grades, requires repeat borrowers to have credit score
    $\geq$ 600, and requires new borrowers to have credit score $\geq$ 640.
    Beginning of Early Auction Regime.\\
    Dec. 19, 2010 & Prosper-rate regime, where Prosper sets rates directly,
    begins, because "auctions made the deployment of funds more time consuming
    with little gain in lender returns,” - Prosper CEO Chris
    Larsen.\footcite{renton}\\
    Aug. 3, 2011 & Last loan in Prosper-rate regime originated.
  \end{tabular}
\label{timeline}
\end{table}

Under all three regimes, lenders could choose to whom to lend and how much to
lend, allowing them to diversify across many loans, thereby decreasing the
variance of their returns. The minimum investment in a loan was \$25, with no
maximum, and lenders could hold as many loans as they wished.
Before a potential borrower's listing is posted on the website, Prosper authenticates their social security number, driver's
license and address, and pulls the borrower's credit history from Experian,
which includes credit score, total number of delinquencies, current
delinquencies, and many other historical credit variables.\footcite{freedman} 
In exchange for verifying borrowers' identities and managing the
marketplace, Prosper charges a 0.5-3 percentage
point origination fee on each loan depending on the borrower's risk profile and
loan duration. 
This fee,usually 1\%, is added to the borrower's monthly payment.
Lenders are also charged a servicing fee of 0.5-1\%. There are no prepayment penalties.
This rate is lower than that charged by most traditional financial institutions, which have the higher costs of physical branches, and pay for human analysis of credit history.\footcite{hanson}
The P2PL model relies on lenders being able to infer the riskiness of borrowers: since Prosper gets fees from monthly payments, they too are incentivized to maximize investor returns, both for short term profit, and long term reputation.

In my sample, Prosper's origination volume, as shown in Figure \ref{fig:origTime}, peaked in the months before the shutdown, and then slowly resumed after the quiet period. Since 2011, when the sample ends, origination volume has eclipsed its 2008 peak.
<<origTime, cache=FALSE, fig.cap='This chart shows two panels. The first panel shows the number of loans funded on Prosper.com in each month of the sample. The second panel shows the total dollars lent to borrowers on Propers.com in each month of the sample.'>>=
g2 = group_by(nd, mgroup, regime) %>%
  summarise(Number_Of_Loans = length(mgroup) 
            ,Dollars_Lent = sum(LoanOriginalAmount)) %>%
    melt(id.vars = c("mgroup",'regime'))

@@

\subsection{Auction Design}
From April 19, 2006 until December 19, 2010. Prosper operated a \emph{variable
rate} model, where lenders and borrowers determined interest rates using a Dutch
auction-like system, where each investor bid consisted of a (dollar amount,
interest rate) pair, such that the dollar amount is greater than \$50, the
minimum bid, and the rate is capped at the borrower's reservation
rate.\footcite{freedman}
In these auctions, borrowers would declare a maximum reservation interest
rate, capped at 36\%, where bidding would start. Over the course of a 7-10 day bidding period, that rate would be bid down by investors, with the loan's interest rate closing at the lowest rate $R_{clearing}$ such that enough investors were willing to fully fund the loan at or below the $R_{clearing}$. All investors who bid at or below the clearing rate would receive the clearing rate. If a borrower could not reach 100\% funding at his reservation rate, the loan would not be funded.

As an illustrative example, suppose Bill the borrower needed to borrow \$100, and set his maximum rate at 20\%.
Let Chris, David and Ellen be investors, with Chris bidding (\$50, 15\%), David bidding (\$50, 18\%) and Ellen bidding (\$30,19\%). In this scenario, Chris and David would win the auction, and each lend lend Bill \$50, and receive an 18\% interest rate. Bill would pay 18\% interest before fees.
\subsection{Early Auction Regime}
In the early auction regime, loans originated between April 19, 2006 and October
15, 2008, auction participants had access to considerable information from
Prosper, including over 60  Borrower characteristics, such as residence,
employment status and an Experian credit score bucket, but few intelligent risk estimates. Additionally, Prosper did not regulate who could apply for a loan on the website. Besides basic fraud prevention, any applicant who received 100\% funding from investors was eligible, for Prosper's first year of operation. On March 1, 2007, amidst poor returns, Prosper announced that listings with credit scores below 520 would no longer be offered to investors. This intervention is not discussed for lack of space, but shows that after a year, more or less, Prosper felt the need to nudge a fully decentralized investment process in the right direction.

@@

\subsection{Prosper-rate Regime}
Under the current \emph{Prosper-rate} model, enacted December 20, 2010, lenders
no longer determine the loan rate in an auction. Instead, Prosper uses the
intelligent credit grades given to investors in the late auction regime to set
interest rates on loans, with slightly less stringent under-writing standards to
determine eligibility. The change was motivated by a desire to accelerate the
listing process, so that investors could reinvest money more quickly. Moreover,
towards the end of the late auction regime, auctions were arriving at interest
rates that were almost completely determined by the credit grade. Under the
fixed-rate regime, investors choose whether or not to invest in a given loan, at
the interest rate set by Prosper, so bids consist only of a dollar amount.

@@

This essay continues Freedman and Jin's critical look at investor decisions in the early
auction regime, but with the benefit of three more years of data, and two more
major Prosper policy changes: informative grading and centralized rate-setting.
Instead of a focus on soft information and social networks, I describe the hard information that
went largely ignored by investors until Prosper aggregated it into more
informative credit grades. Instead of investor learning as the catalyst that led
to a more efficient market, I frame Prosper's rule changes as catalysts that
force investors to make better decisions.

Inconsistent with Hayek, the data suggest that when investors are left on their own
to set prices, markets are not nearly as efficient as the prices set by ``a
single mind'', namely a market maker who aggregates the data. Furthermore, as interest rate setting becomes more centralized, interest rates become more sensitive to the hard credit variables that best predict default.

\section{Data}
My sample consists of 39,555 three year, unsecured (no collateral) Loans, worth
\$231.7 Million,  originated from 2006-2011. The data was downloaded off Prosper.com.
Borrowers apply for loans by creating a public listing on the Prosper.com website, and can request anywhere from \$1,000 to \$35,000.\footnote{The minimum increased to \$2,000 effective Nov. 1, 2011}.
Since our analysis requires measurement of loan performance, which can only take
place three  years after origination (given 3 year maturity on all loans), the
sample is ended in 2011, as we do not know what will end up happening to more recently originated loans. 
As shown in Table \ref{sample}, this leaves us with many more early auction regime loans than the post shutdown regimes.
<<sampTab, cache=FALSE>>=
samp.tab = mf(count(nd,regime))
colnames(samp.tab) = c('Regime','n','Freq.')
xtable(samp.tab,kdigits=4, caption='Observations per regime. Each observation is
one funded loan.', label='sample')
@

@@

\subsection{Measuring Loan Performance with ROI}
<<roivstatus,cache=FALSE,fig.height=3,fig.cap="This chart shows the distribution of ROI for completed and defaulted loans over all three regimes. A continuous ROI variable offers more information about the quality of an investment than a binary default indicator. Dotted lines indicate mean ROI for category.",fig.height=3,fig.pos='H'>>=
nd$d.factor =  ifelse(nd$defaulted == 1, 'Defaulted','Completed')
smeans = summarise(group_by(nd,d.factor), mRoi = mean(roi, na.rm=TRUE))
ggplot(nd, aes(x=roi,fill=d.factor)) +
  geom_histogram(binwidth=.01)+ 
  geom_vline(data=smeans, aes(xintercept=mRoi,color=d.factor),linetype='longdash') + 
  theme_bw() +
  labs(title='Distribution of ROI for each Loan Outcome (all regimes)', x='ROI',y='Count',fill='Loan Status') +
  scale_x_continuous(limits=c(-1, 1), breaks=seq(-1,1,by=.25),label=percent) +
  facet_wrap(~d.factor,ncol=1,scales='free_y')
  @
